---
title: "Hands-on Exercise 2: Thematic Mapping and Geovisualisation with R"
author: "Victoria Grace ANN"
execute: 
  warning: false
date: "6 January, 2024"
date-modified: "last-modified"
---

# Introduction

Thematic mapping involves the use of map symbols to visualise selected properties of geographic features that are not naturally visible, such as population, temperature, crime rates and property prices.

Geovisualisation works by providing graphical ideation to render a place, phenomenon or a process.

# Setting up

## Data 

- Master Plan 2014 Subzone Boundary (Web) from [data.gov.sg ](https://data.gov.sg)
- Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2011-2020 in csv format from [Department of Statistics, Singapore] (https://singstat.gov.sg)

## Package

*tmap* package is used for this exercise. To install these packages in RStudio:

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## Importing Geospatial Data into RStudio

```{r}
mpsz <- st_read(dsn = "data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
```

Examine the `mpsz` content by parsing:

```{r}
mpsz
```

## Importing Attribute Data into RStudio

After downloading the *respopagessex2011to2020.csv* file, save the file into an R dataframe, *popdata*.

Access the csv file using *read_csv()* in the **readr** package:

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")

show_col_types = FALSE
```

# Data preparation

Focusing on year 2020, prepare a data table.It should include variables PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY.

-   YOUNG: age group 0 to 4 until age group 20 to 24

-   ECONOMY ACTIVE: age group 25-29 until age group 60-64

-   AGED: age group 65 and above

-   TOTAL: all age groups, and

-   DEPENDENCY: the ratio between young and aged against economy active group.

## Data wrangling

These are the following wrangling and transformation functions that will be used: - *pivot_wider()* from **tidyr** package - *mutate(), filter(), group_by()* and *select()* from **dplyr** package

*Note: Added additional argument, .groups, in the summarise() function. Error occurred initially.*

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`), .groups = "drop") %>%
  ungroup() %>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

## Joining attribute data and geospatial data

Before performing the georelational join, the values in the PA and SZ fields need an uppercase conversion. The SUBZONE_N and PLN_AREA_N are already in uppercase.

```{r}
popdata2020 <- popdata2020 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Next, *left_join()* of **dplyr** is used to join the geographical data and attribute table using planning subzone name e.g. *SUBZONE_N* and *SZ* as the common identifier.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

-   The *left_join()* from **dplyr** package is used with `mpsz` simple feature data frame as the left data table is to ensure that the output will be a simple feature dataframe.

Save the changes.

```{r}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")
```

-   RDS files take less space than textual data formats like CSV, JSON. Further, RDS also reads and writes faster than those.

# Choropleth Mapping Geospatial Data Using *tmap*

Choropleth mapping is the symbolisation of enumeration units, including countries, provinces, states, or census units. This involves the use of area patterns or graduated colours.

**Scenario** A social scientist may need to use a choropleth map to portray the spatial distribution of aged population of Singapore w.r.t the Master Plan 2014 Subzone Boundary.

There are two approaches to prepare thematic maps using *tmap*: 1. Plotting a quick thematic map using *qtm()* 2. Plotting a customisable thematic map using *tmap* elements

## Method 1: Plotting a quick choropleth map using *qtm()*

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")
```

-   *tmap_mode()* with "plot" option is used to produce a static map. For interactive mode, "view" option should be used.
-   *fill* argument is used to map the attribute (i.e. DEPENDENCY)

## Method 2: Creating a choropleth map using *tmap's* elements

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "Dependency ratio") +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

The following subsection explains the *tmap* functions used to plot these elements.

### Drawing a base map

The basic building block of **tmap** is *tm_shape* followed by one or more layer elements such as *tm_fill()* or *tm_polygons()*.

*tm_shape()* is used to define the input data *mpsz_pop2020* and *tm_polygons()* is used to draw the planning subzone polygons

```{r}
tm_shape(mpsz_pop2020) + tm_polygons()
```

### Drawing a choropleth map using *tm_polygons()*

```{r}
tm_shape(mpsz_pop2020) + tm_polygons("DEPENDENCY")
```

-   The default interval bin is called "pretty". A detailed discussion of the classification methods supported by **tmap** will be provided in a later sub-section.
-   The default colour scheme used is `YlOrRd` of ColorBrewer. More about the colour scheme can be learnt in a later section.

### Drawing a choropleth map using *tm_fill()* and *tm_border()*

*tm_polygons()* is a wrapper of *tm_fill()* and *tm_border()*. *tm_fill()* shades the polygons using the default colour scheme and *tm_borders()* adds the borders of the shapefile onto the choropleth map.

Draw the choropleth map using *tm_fill()* alone,

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```

-   The planning subzones are shared according to the respective dependency values.

*tm_boarders()* is used to draw up the boundaries of the subzones:

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY") +
  tm_borders(lwd = 0.1,  alpha = 1)
```

The alpha argument is used to define transparency number between 0 (totally transparent) and 1 (not transparent). By default, the alpha value of the col is used (normally 1).

Beside alpha argument, there are three other arguments for tm_borders(), they are:

-   col = border colour,
-   lwd = border line width. The default is 1, and
-   lty = border line type. The default is “solid”.

## Data classification methods of *tmap*

Classification helps to group a large number of observations into data ranges or classes.

**tmap** provides a total of ten data classification methods (i.e. *fixed, sd, equal, pretty (default), quantile, kmeans, hclust, bclust, fisher, jenks*).

To define a data classification method, the *style* argument of *tm_fill()* or *tm_polygons()* will be used.

### Plotting choropleth maps with built-in classification

Within the **tm_fill()** argument, the number of classes can be customised. First we will use the **jenks** method, which is also known as natural breaks, is based on natural groupings inherent in the data. Data is clustered into groups that minimise the within-group variance and maximises the between-group variance.

The following code chunk depicts a quantile data classification that uses 5 classes.

#### jenks
```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n=5,
          style="jenks") +
  tm_borders(alpha=0.5)

```

#### equal
Next, we will try the **equal** data classification method.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n=5,
          style="equal") +
  tm_borders(alpha=0.5)

```
-   The distribution of quantile data classification is more evenly distributed than the equal data classification method.


**tmap** actually has 10 data classification methods:
- jenks (or natural breaks as mentioned)
- equal (as mentioned)
- fixed
- sd
- pretty (default)
- quantile
- kmeans
- fisher
- hclust
- bclust

Looking closer at...
#### sd
```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n=5,
          style="sd") +
  tm_borders(alpha=0.5)

```
#### kmeans 
```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n=5,
          style="kmeans") +
  tm_borders(alpha=0.5)

```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n=10,
          style="kmeans") +
  tm_borders(alpha=0.5)

```

#### fisher
```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n=5,
          style="fisher") +
  tm_borders(alpha=0.5)

```
- At a glance, using Fisher and KMeans lead to similar visualisations.

#### hclust
**hclust** is hierarchical clustering. It is a method aiming to create a hierarchy of clusters based on their similarity. Each data point starts as an individual cluster and then progressively merges or splits clusters until a stopping criterion is met. 
```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n=5,
          style="hclust") +
  tm_borders(alpha=0.5)

```
```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n=20,
          style="hclust") +
  tm_borders(alpha=0.5)

```

#### bclust
**bclust** is bagged clustering which creates multiple subsets of the original dataset through resampling. Each subset is then used to train an individual clustering model, and the final cluster assignments are obtained by combining the results from all models.
```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n=5,
          style="bclust") +
  tm_borders(alpha=0.5)

```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n=4,
          style="bclust") +
  tm_borders(alpha=0.5)

```
```

### Plotting choropleth maps with custom breaks
Breakpoints can be set explicitly to override the default category breaks. This can be done with the help of the *breaks* argument in *tm_fill()*. For n categories, n+1 elements must be included in the *breaks*.

Get descriptive statistics before setting the break points.
```{r}
summary(mpsz_pop2020$DEPENDENCY)

```

Looking at the summary statistics, the break point can be set to 0.60, 0.70, 0.80, and 0.90. The minimum and maximum breaks must also be included, which are 0 adn 100 respectively. These would translate to the *breaks* vector, `c(0, 0.60, 0.70, 0.80, 0.90, 1.00)`.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(alpha = 0.5)
```

## Colour Scheme
**tmap** also supports colour ramps. They can be defined by the analyst or preselected from the **RColorBrewer** package.

### ColorBrewer palette
The colour of the choropleth map can be reassigned to a colour of our preference. Simply add the *palette* argument into *tm_fill* as shown in the code chunk below.
```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = "quantile",
          palette = "Blues") +
  tm_borders(alpha = 0.5)
```

To reverse the colour shading, add a "-" prefix.
```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "-Greens") +
  tm_borders(alpha = 0.5)
```
- The colour scheme has been reversed as depicted by the decreasing colour intensity in the legend.

## Map Layouts
Map layout refers to the combination of all map elements into a comprehensive map. In maps, it is crucial to include their elements -- title, scale bar, compass, margins and aspect ratios.

### Map Legend
In **tmap**, the *legend* options can be customised to change the placement, format and its appearance.
```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "jenks", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone \n(Jenks classification)",
            main.title.position = "center",
            main.title.size = 1,
            legend.height = 0.45, 
            legend.width = 0.35,
            legend.outside = FALSE,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

### Map Style
The layout of the map can also be adjusted using *tmap_style()*.

#### Classic 
```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

### Cartographic Furniture
**tmap** also provides arguments to draw other important map elements like compass, scale bar and grid lines.

To add compass, scale and gridlines, pay attention to how *tm_compass()*, *tm_scale_bar()* and *tm_grid()* are used.
```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "No. of persons") +
  tm_layout(main.title = "Distribution of Dependency Ratio \n by Planning Subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authority (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

## Drawing Small Multiple Choropleth Maps
Small multiple maps, also referred as **facet maps**, comprise of many adjacent maps. These facets enable easier visualisation of how spatial relationships change with respect to another variable such as time.

In **tmap**, small multiple maps can be plotted in three ways:
1. Assigning multiple values to at least one of the aesthetic arguments
2. Defining a group-by variable in *tm_facets()*, and
3. Creating multiple stand-alone maps with *tmap_arrange()*

### Assigning multiple values to at least one of the aesthetic arguments

Small multiple choropleth maps can be created by defining the number of columns, **ncols**, in **tm_fill()**.
```{r}
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal", 
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

Multiple values can be assigned to aesthetic arguments like style and palette. The visual output of the maps will be according to the corresponding order of the argument values.
```{r}
tm_shape(mpsz_pop2020)+ 
  tm_polygons(c("DEPENDENCY","AGED"),
          style = c("equal", "quantile"), 
          palette = list("Blues","Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```

### Defining a group-by variable in *tm_facets()*
*tm_facets()* can help to group categorical data like regions and subzone areas such that the generated facet maps will zoom in to the specified variable.
```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) + 
  tm_facets(by="REGION_N", 
            free.coords=TRUE, 
            drop.shapes=TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

### Creating multiple stand-alone maps with *tmap_arrange()*
The order of the maps can be reorganised by assigning each map to a unique variable then ordering the variables using *tmap_arrange()*.
```{r}
youngmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("YOUNG", 
              style = "quantile", 
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("AGED", 
              style = "quantile", 
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```
- *asp* refers to th aspect ratio of the maps. Its default value is 0.

## Mapping Spatial Objects Meeting a Selection Criterion
Map outputs can be targeted by using selection functions to meet the selection criterion
```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45, 
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```
