---
title: "In-class Exercise 5"
author: "Victoria Grace ANN"
execute: 
  warning: false
  eval: true
  echo: true
format:  
  html: 
    code-summary: "Show the code"
    toc-depth: 4
date: "5 February, 2024"
date-modified: "last-modified"
editor: visual
---

# Installing and Loading Packages

-   [**sfdep**](https://sfdep.josiahparry.com/) creates a sf and tidyverse friendly interface to the package.

```{r}
pacman::p_load(sf, tmap, sfdep, tidyverse, dplyr)
```

# Data

The two data sets we will use are: - Hunan, a geospatial data in ESRI shape file format - Hunan_2012, aspatial attribute data in csv format

## Import geospatial data

```{r}
hunan <- st_read(dsn="data/geospatial", layer="Hunan")
```

## Importing aspatial data

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### Combining aspatial fields to geospatial data

```{r}
hunan_GDPPC <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

### Plotting Choropleth Map

```{r}
tmap_mode("plot")
tm_shape(hunan_GDPPC) +
  tm_fill(col="GDPPC",
          style = "quantile",
          palette = "Blues",
          title = "GDPPC") +
  tm_layout(main.title = "Distribution of GDP per capita by County",
            main.title.position = "center",
            main.title.size =1.2,
            legend.height = 0.45,
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha = 0.2)
```

# Global Measures of Spatial Association

The queen method is used to derive the contiguity weights

```{r}
wm_q <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb,
                         style = "W"),
         .before = 1)
```

## Computing Global Moran's I

```{r}
moranI <- global_moran(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)
glimpse(moranI)
```

```{r}
global_moran_test(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)
```

### Performing Global Moran's I permutation test

It is better to conduct permutation tests.

```{r}
set.seed(1234)
```

```{r}
global_moran_perm(wm_q$GDPPC,
                  wm_q$nb,
                  wm_q$wt,
                  nsim = 99) 
```

-   `nsim` refers to the number of simulations
-   Since the p-value is \< 0.05, there is sufficient evidence to reject the null hypothesis that the spatial distribution of GDP per capita resembles a random distribution, or rather the distribution is independent from spatial. Since Moran's I statistic is \> 0, we can infer that the spatial distribution shows signs of clustering.
